---
title: "Use ServCat Files in a RAG Store"
description: "Search ServCat, download topic-specific files, and use them as source documents for a RAG workflow."
format:
  html:
    toc: true
    toc-depth: 2
execute:
  eval: false
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This article shows a workflow for using `servcat` as a source of documents for a retrieval-augmented generation, or RAG, workflow.

A RAG store is a searchable collection of source documents used by a large language model. Instead of asking the model to answer only from its general training data, a RAG application first retrieves relevant text from your document store and then asks the model to answer using that retrieved material.

ServCat is useful for this because it contains authoritative U.S. Fish and Wildlife Service documents and metadata. A topic-specific ServCat corpus can help an LLM answer questions about a refuge, species, monitoring project, recovery plan, or other Service topic while keeping the answer grounded in source documents.

The basic workflow is:

1. Search ServCat for a topic.
2. Choose the references you want.
3. Review the files attached to those references.
4. Download the files.
5. Keep a manifest so each local file can be traced back to ServCat.
6. Optionally load the files into a RAG store with packages such as `ragnar` and `ellmer`.

## Packages

```{r}
#| label: packages

library(servcat)
library(dplyr)
library(purrr)
library(readr)
library(fs)
library(stringr)
```

Set `secure = TRUE` only when you are authorized to use the secure ServCat API and have configured the required API key.

```{r}
#| label: secure

secure <- FALSE
```

## Search for references

Start with a topic. This example searches for public ServCat references about Kodiak mountain goats that have digital files.

```{r}
#| label: search-topic

topic <- "Kodiak mountain goat"

criteria <- list(
  quickSearch = topic,
  visibility = "public",
  digitalResources = list(
    list(
      order = 1,
      logicOperator = NULL,
      type = "DigitalFile",
      fieldName = NULL,
      searchText = NULL
    )
  )
)

refs <- search_references(
  criteria = criteria,
  top = 50,
  page = 1,
  orderby = "dateOfIssue",
  sort = "DESC",
  secure = secure
)

refs
```

Review the results and choose the references you want to include.

```{r}
#| label: choose-references

candidate_refs <- refs |>
  distinct(referenceId, .keep_all = TRUE) |>
  select(any_of(c(
    "referenceId",
    "title",
    "referenceType",
    "dateOfIssue",
    "displayCitation"
  )))

candidate_refs

reference_ids <- candidate_refs |>
  slice_head(n = 10) |>
  pull(referenceId)

reference_ids
```

For a real RAG store, it is usually better to review the results and set `reference_ids` manually or by evaluating the content of each file to verify that it in fact directly relates to the topic.

## Get file metadata

Use `get_files()` to list the files attached to each reference.

```{r}
#| label: get-files

files <- map(reference_ids, \(id) {
  get_files(reference_id = id, secure = secure) |>
    mutate(referenceId = id)
}) |>
  list_rbind() |>
  mutate(extension = str_to_lower(path_ext(fileName)))

files |>
  select(any_of(c(
    "referenceId",
    "resourceId",
    "fileName",
    "description",
    "fileSize_kb",
    "extension",
    "downloadLink"
  )))
```

In this example, we focus on common document formats that are likely to contain extractable text. We will keep Word documents, PDFs, HTML files, plain text files, and PowerPoint presentations.

```{r}
#| label: select-files

rag_extensions <- c(
  "doc", "docx",
  "pdf",
  "html", "htm",
  "txt",
  "ppt", "pptx"
)

selected_files <- files |>
  filter(extension %in% rag_extensions)

selected_files |>
  select(any_of(c(
    "referenceId",
    "resourceId",
    "fileName",
    "extension",
    "fileSize_kb",
    "downloadLink"
  )))
```

This review step matters. Remove files that are not relevant, duplicated, restricted, too large, or unlikely to convert cleanly to text. PowerPoint files can work well when they contain textual slide content, but image-heavy presentations may need additional processing.

## Download files

Create a local folder for the downloaded files.

```{r}
#| label: download-folder

topic_slug <- topic |>
  str_to_lower() |>
  str_replace_all("[^a-z0-9]+", "-") |>
  str_replace_all("(^-|-$)", "")

download_dir <- path("servcat-files", topic_slug)
dir_create(download_dir)
```

Download the selected files. `download_files()` works one reference at a time, so this code groups the selected resource IDs by reference. If you have a lot of files, this could take some time.

```{r}
#| label: download-selected-files

download_plan <- selected_files |>
  group_by(referenceId) |>
  summarise(
    resource_ids = list(resourceId),
    .groups = "drop"
  )

downloads <- download_plan |>
  mutate(
    result = map2(
      referenceId,
      resource_ids,
      \(reference_id, resource_ids) {
        download_files(
          reference_id = reference_id,
          resource_ids = resource_ids,
          path = download_dir,
          secure = secure,
          overwrite = FALSE
        )
      }
    )
  ) |>
  pull(result) |>
  list_rbind()

downloads
```

Check for failed downloads before continuing.

```{r}
#| label: check-downloads

downloads |>
  count(success)

downloads |>
  filter(!success | is.na(success)) |>
  select(any_of(c("referenceId", "resourceId", "fileName", "error")))
```

## Save a manifest

A manifest records where each local file came from. Keep it with the downloaded files and with any RAG store you build from them.

```{r}
#| label: save-manifest

ref_summaries <- get_reference_summaries(
  reference_ids = reference_ids,
  secure = secure
)

manifest <- downloads |>
  left_join(
    files |>
      select(any_of(c(
        "referenceId",
        "resourceId",
        "description",
        "fileSize_kb",
        "extension",
        "downloadLink"
      ))),
    by = c("referenceId", "resourceId")
  ) |>
  left_join(
    ref_summaries |>
      select(any_of(c(
        "referenceId",
        "title",
        "referenceType",
        "dateOfIssue",
        "displayCitation"
      ))),
    by = "referenceId"
  )

manifest_path <- path(download_dir, "servcat-download-manifest.csv")
write_csv(manifest, manifest_path)

manifest_path
```

The manifest should include, when available, the ServCat `referenceId`, `resourceId`, original `fileName`, local file path, download link, title, citation, and download status.

## Optional: create a RAG store with `ragnar`

The downloaded files are now ready to use as source documents. The example below shows the general pattern with `ragnar`: create a store, convert files to Markdown, split them into chunks, insert the chunks, and build an index.

This section can call an embedding API, so make sure your model-provider API key is configured before running it.

```{r}
#| label: ragnar

library(ragnar)

store_path <- path(download_dir, "servcat-rag.duckdb")

store <- ragnar_store_create(
  location = store_path,
  embed = \(x) ragnar::embed_azure_openai(x, 
                                          model = "text-embedding-3-small", 
                                          endpoint = "https://api-dev.ai.doi.net/",
                                          api_version = "2024-02-15-preview"),
  overwrite = TRUE,
  name = "servcat_files",
  title = "ServCat files"
)

successful_downloads <- manifest |>
  filter(success, !is.na(localPath), file_exists(localPath)) |>
  mutate(
    origin = paste0(
      "servcat://reference/", referenceId,
      "/resource/", resourceId,
      " | ", fileName
    )
  )

for (i in seq_len(nrow(successful_downloads))) {
  message("ingesting: ", successful_downloads$fileName[[i]])
  chunks <- successful_downloads$localPath[[i]] |>
    read_as_markdown(origin = successful_downloads$origin[[i]]) |>
    markdown_chunk()

  ragnar_store_insert(store, chunks)
}

ragnar_store_build_index(store)
```

Test retrieval before connecting the store to a chat model.

```{r}
#| label: retrieve

store <- ragnar_store_connect(
  location = store_path,
  read_only = TRUE
)

ragnar_retrieve(
  store = store,
  text = "What information is available about Kodiak mountain goat monitoring?",
  top_k = 5
)
```

## Optional: use the store with `ellmer`

After retrieval works, you can register the store as a tool for an `ellmer` chat object.

```{r}
#| label: ellmer

library(ellmer)

chat <- ellmer::chat_azure_openai(
    model = "gpt-4o",
  endpoint = "https://api-dev.ai.doi.net/",
  params = ellmer::params(temperature = 0.1),
  system_prompt = paste(
    "You answer questions using a local knowledge store built from ServCat files.",
    "Retrieve relevant passages before answering.",
    "Use the retrieved passages as the source of truth.",
    "When possible, cite ServCat reference IDs and resource IDs.",
    sep = "\n"
    )
  )

ragnar_register_tool_retrieve(
  chat = chat,
  store = store,
  store_description = "ServCat files downloaded for the selected topic",
  top_k = 10
)

chat$chat("Summarize the ServCat information about Kodiak mountain goat monitoring.")
```

## Tips

Start with a small, curated set of references. A smaller store with relevant files usually works better than a large store with mixed or duplicated material.

Keep the manifest. It is the link between model answers, local files, and ServCat source records.

Review access and use constraints before putting documents into a shared RAG application.

Inspect text extraction quality. Scanned PDFs, image-heavy reports, and image-heavy PowerPoint files may need additional processing before they work well in a RAG store.

## Summary

Use `search_references()` to find topic-specific ServCat records, `get_files()` to inspect attached files, and `download_files()` to download selected Word, PDF, HTML, text, and PowerPoint resources. Keep a manifest with ServCat identifiers and local file paths. The downloaded files can then be converted, chunked, embedded, and indexed with RAG tools such as `ragnar`, then used from an LLM interface such as `ellmer`.
